---
title: "Bumpus's sparrows"
---

```{r, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4.5, dpi = 96)
```

Bumpus (1899) measured 136 house sparrows brought in after a winter storm in Providence, Rhode Island, of which 72 survived. The data ship with the package as `bumpus`, with nine traits. Janzen and Stern (1998) reanalysed them by sex with logistic regression, and their tables are the reference the package is checked against.

## Gradients on all nine traits

```{r}
library(Lande)
traits <- c("total_length", "wingspread", "weight", "head_length", "humerus",
            "femur", "tibiotarsus", "skull_width", "sternum")
report <- selection_report(bumpus, "survival", traits, fitness_type = "binary")
report[report$Type == "Linear", ]
```

Survival selected against total length and weight and for sternum length, with the other traits held constant. The p-values come from the logistic model; the coefficients and their standard errors from least squares on relative fitness, so they are on the Lande and Arnold scale.

## Within each sex

Janzen and Stern also standardised within sex and estimated each sex separately, on log-transformed traits. On logged traits the package gives their gradients to three decimals; on the raw traits used here the numbers differ a little.

```{r}
by_sex <- selection_coefficients(bumpus, "survival", traits, fitness_type = "binary",
                                 group = "sex", return_grouped = TRUE)
by_sex[by_sex$Type == "Linear" & by_sex$Term %in% c("total_length", "weight"), ]
```

Males (87 birds) carry the selection against total length; females (49) the selection against weight, with wide standard errors.

## The fitness function for total length

```{r}
prep <- prepare_selection_data(bumpus, "survival", "total_length")
fit <- univariate_spline(prep, "survival", "total_length", k = 6, bootstrap = TRUE, n_boot = 200)
plot_univariate_fitness(fit, "total_length")
```

With the default basis size, k = 10, the UBRE criterion settles on a wiggly curve (edf 8.1) for these 136 birds; k = 6, or REML with k from 4 to 10, gives a smooth one (edf 2.2 to 2.5). Survival is highest a little below average length and falls for longer birds. The band is a percentile bootstrap of the spline.

## Checks

```{r}
check_selection_assumptions(bumpus, "survival", c("total_length", "weight", "humerus"))
```

With nine traits the quadratic model has 54 terms for 64 deaths, so only the linear gradients in the report above are worth reading.
